D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-21 cuttingroom.tex
Drop "Laws of Physics" section
But it lives on in cuttingroom.tex.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

>>>>

\subsection{Laws of Physics}
\label{sec:Laws of Physics}

Evidence to date suggests that the universe is
causal~\cite{Plato360BC-causality},
the speed of light is finite~\cite{OleRoemer1671SpeedOfLight}, and
atoms are of nonzero size~\cite{JeanBaptistePerrin1923AtomSize}.
These fundamental laws of physics are empirical in nature, but
are backed by a great many observations extending back to a time
preceding electronic computers, let alone computer languages that
support concurrency.

The relationship between these three laws of physics and OOTA
cycles is described in the following sections.

\subsubsection{Causality}
\label{sec:Causality}

Plato's ``Timaeus'' notwithstanding, there has been much recent debate
as to whether the universe is in fact causal with cause necessarily
preceding effect.
However, no one has managed to construct a real time-travel machine
or any other apparatus that might result in effect preceding cause.
% as discussed at length in Appendix~\ref{sec:Debates on Causality} and 
% the URL that it provides.

This paper will therefore assume that effect cannot precede cause in
any real-world computing system.

\subsubsection{Finite Speed of Light}
\label{sec:Finite Speed of Light}

If the speed of light was infinite, then information could be transmitted
at infinite speed, arriving at any destination, no matter how remote,
in zero time.
This situation would permit a load to return the value stored by some
other thread at the exact instant that the store executed, which
would in turn remove propagation delay as a factor preventing
causal loops such as OOTA cycles from forming.

However, all macroscopic measurements to date have shown the speed of
light to be finite.
This paper will therefore assume that the speed of light is finite, and
that the maximum speed at which information can propagate is also finite.

As long as empirical evidence continues to validate this assumption, it
is not possible to form causal loops based on infinite-speed propagation
of information.

\subsubsection{Non-Zero Sized Atoms}
\label{sec:Non-Zero Sized Atoms}

If atoms were of zero size, a zero-sized computer might be constructed.
In such a computer, even given finite speed of light, information could
propagate among computing elements in zero time.
This situation would again permit a load to return the value stored by
some other thread at the exact instant that the store executed, which
would again in turn remove propagation delay as a factor preventing
causal loops such as OOTA cycles from forming.

However, there has to date been no successful attempt to construct a
zero-sized computing device capable of running nontrivial C++ programs.

This paper will therefore assume that any computing device capable of
running C++ will be of nonzero size, which rules out formation of
causal loops based on zero-size computing systems.

\subsubsection{Laws of Physics: Consequences}
\label{sec:Laws of Physics: Consequences}

These three laws of physics mean that if one thread loads the value
stored by some other thread, that load must have occurred later
in global time than did the store.
In other words, the reads-from relation is temporal in nature.

In contrast, the C++11 modification-order relation is atemporal,
because computing systems can and observably do determine the
modification order of a set of concurrent stores long after the fact
(see Appendix~\ref{app:Inter-Process Communications}).
This means that the store that executed earlier in global time
might appear later in the stored-to object's modification
order~\cite{McKenney20xxParallelProgramming}.
The atemporal nature of C++11 modification order is due to the hardware
store-buffer optimizations used by modern multicore systems.

This atemporality is also why cycles are not considered to have
OOTA cycles when at least one link from one thread to the next
involves both threads storing to the same object.
Similarly, cycles having at least link where one thread loads from an
object and the next thread stores to that object are also not considered
to be OOTA cycles.

Current programming-language abstract machines do not capture these
temporal constraints, which makes it more difficult for them to
efficiently rule out OOTA behavior.
